Comparison of MCMC algorithms in Stochastic Volatility Models

### 2023-08-21 writeups/research_proposal/propsal.qmd
fixed figures

---
title: "Comparison of MCMC algorithms in Stochastic Volatility Models"
format: pdf
---

@@

# Results
SBC for the HMC algorithm and the KSC Gaussian approxmiation are conducted with simulated dataset sizes of 1000 observations and 1000 simulation interations. The preliminary results focus on a subset of parameters since there are parameters for each latent state (i.e one for each data observation) plus the static parameters.

@@

![Distribution of parameter ranks for HMC algorithm](../../simulation_output/sbc_ncp_ksc_priors_0.999_adapt_delta_premade_datasets_r7_1000_iterations/static_state_hist.png){width=75%}

@@

Figure 3 displays the results from KSC's Gaussian approximation. Compared to HMC, the distribution of ranks deviate heavily from the expected discrete uniform distribution. The lack of calibration relative to the HMC algorithm is expected. The Gaussian Mixture Approximiation only approximates the log chi squared distribution. The data generating process is from the specified Stochastic Volatility model. We expect that since the HMC strategy samples directly from the true data generating process that it would have more success at

![Distribution of parameter ranks for HMC algorithm (5000 iterations)](../../simulation_output/sbc_cp_ksc_model_cp_dgf_r1/static_state_hist.png){width=75%}

# Conclusion and future research considerations

## Plan
The KSC model also also apply a Metropois Hastings correction for the fact that the sampling strategy approximates the error distribution.

Extensions to the model table:

    + Stan (HMC), KSC (GMA), KSC (GMA + MH correction)
    + CP, NCP

Parameterisation

## Additional considerations
- Out of sample prediction on real dataset

>>>>

---
title: "Comparison of MCMC algorithms in Stochastic Volatility Models"
format:
  pdf:
    fig-pos: H
---

@@

# Results
SBC for the HMC algorithm and the KSC Gaussian approxmiation are conducted with simulated dataset sizes of 1000 observations, 1000 simulation interations with post warmup/burnin samples of 999 (which gives 1000 possible rank statistics). The histogram is drawn with 20 bins and 50 observations in each bin which is the expected count for a discrete uniform distribution. The preliminary results focus on a subset of parameters since there are parameters for each latent state (i.e one for each data observation) plus the static parameters.

@@

![Distribution of parameter ranks for HMC algorithm (1000 iterations)](../../simulation_output/sbc_ncp_ksc_priors_0.999_adapt_delta_premade_datasets_r7_1000_iterations/static_state_hist.png){width=75%}

@@

Figure 3 displays the results from KSC's Gaussian approximation. Compared to HMC, the distribution of ranks deviate heavily from the expected discrete uniform distribution. The lack of calibration relative to the HMC algorithm is expected. The Gaussian Mixture Approximiation only approximates the log chi squared distribution. The HMC strategy samples directly from the same likelihood as the data generating process.

![Distribution of parameter ranks for KSC sampling strategy (1000 iterations)](../../simulation_output/sbc_cp_ksc_model_cp_dgf_r1/static_state_hist.png){width=75%}

An alternative to checking for uniformity of all the parameters is to calculate the chi squared statistics for the counts in each bin. Let $b_j$ be the number of counts and $e_j$ the expected count in bin $j$. Then the chi squared statistic is given by:

$$
\begin{aligned}
\chi^2 = \sum_{j=1}^J \frac{(b_{j} - e_{j})^2}{e_j}
\end{aligned}
$$

These chi squared statistics can be considered independent draws from a chi squared distribution with 19 degrees of freedom (since there are 20 bins). Visualising this should give a chi squared distribution (if performing a hypothesis test, this would be the sampling distribution under a null of uniformity).

Figure 4 and 5 are the chi squared statistic histograms results of the HMC algorithm and KSC algorithm respectively. The vertical black line marks the 95th quantile which represents the critical value for hypothesis testing. The HMC chi squared statistics look like they are sampled from the correct distribution whereas KSC are inconsistent with the expected distribution.

![Chi squared statistics for HMC](../../simulation_output/sbc_ncp_ksc_priors_0.999_adapt_delta_premade_datasets_r7_1000_iterations/chi_sq_hist.png){width=70%}

![Chi squared statistics for KSC](../../simulation_output/sbc_cp_ksc_model_cp_dgf_r1/chi_sq_hist.png){width=70%}

# Conclusion and next steps
KSC also also apply a Metropolis Hastings reweighting to correct the approximation of the error distribution. According to the authors, the correction ensures that samples are from the exact posterior density. The next step is to perform this correction and evaluate the SBC results.

Another factor that may affect sampling performance is the parameterisation of the model. This can cause significant problems to samplers in models with increasing complexity. A classic example of this is Neal's funnel which arises from Hierarchical models - causing bias and performance issues with the HMC sampler. The solution is to rewrite the model with non centered parameterisation which significantly improves the performance of the sampler. Another step in this research is to consider the results of SBC under different model parameterisations. The models and parametiersations for this research are summarised in the below table:

![Model matrix](model_table.png)
